# Setup

There are countless number of guides to setting up your local R and RStudio environment. If you're learning on your own, its easy to get some of the below applications and packages installed.

## R and RStudio

One of the most respected introductions to R is [R for Data Science](https://r4ds.had.co.nz/introduction.html) by Hadley and Garrett, and the [Prerequisities](https://r4ds.had.co.nz/introduction.html#prerequisites) section is set up for installing R. 

## Common packages

We will use several common packages; if you haven't installed the yet, feel free to uncomment the installation line and load them up.

```{r, eval = FALSE}
install.packages(c('tidyverse'))
```

## The AUDITPACKAGENAME

```{r}
# TODO publish audit package
```

## Github (or another git-related brand)

There are many different companies that provide git - Azure, Atlassian, Gitlab, etc. If you're learning, github is a fantastic free source to set up shop at. Within your own company, see what the IT and developers are using, as its far more convenient to jump onto that.

Like the R tutorial, [Happy Git and GitHub for the useR](https://happygitwithr.com) by Jenny Bryan et. al. is a practical guide to installing R, setting up keys and the core fundamentals.
