# (PART) Technical {-}

# Setup 

The next set of technical chapters are intended to give you more hands-on guidance for using R in an audit environment. You should have R and RStudio installed, as well as some of the packages we intend to use.

There are countless number of guides to setting up your local R and RStudio environment. If you're learning on your own, its easy to get some of the below applications and packages installed.

## R with RStudio {#setup-r}

One of the most respected introductions to R is [R for Data Science](https://r4ds.had.co.nz/introduction.html) by Hadley Wickham and Garrett Grolemund, and the [Prerequisities](https://r4ds.had.co.nz/introduction.html#prerequisites) section is set up for installing R. 

## Common packages

We will use several common packages; if you haven't installed them yet, feel free to install them into R:

```{r, eval = FALSE}
install.packages(c('tidyverse', 'lubridate', 'httr', 'jsonlite', 'DBI', 'devtools'))
```

## Audit Package helpers

For those customers who use the cloud-based Galvanize Highbond (and specifically the Results Module), we will use the R package [galvanizer](https://github.com/jonlinca/galvanizer) to use Results information in the applied analytics components.

As these packages are in active development, they are not currently available on CRAN. You are able to install any package on GitHub - simply install [RTools](https://cran.r-project.org/bin/windows/Rtools/) on your computer, which allows you to bundle your own packages. 

```{r, eval = FALSE}
remotes::install_github('jonlinca/galvanizer')
```

## GitHub (or another git-related brand)

There are many different companies that provide git - Azure, Atlassian, Gitlab, etc. If you're learning, GitHub is a fantastic free source to set up shop at. Within your own company, see what the IT and developers are using, as its far more convenient to jump onto that.

Like the R tutorial, [Happy Git and GitHub for the useR](https://happygitwithr.com) by Jenny Bryan et. al. is a practical guide to installing R, setting up keys and the core fundamentals.
