# (PART) Technical {-}

# Setup 

The next set of technical chapters are intended to give you more hands-on guidance for using R in an audit environment. You should have R and RStudio installed, as well as some of the packages we intend to use.

There are countless number of guides to setting up your local R and RStudio environment. If you're learning on your own, its easy to get some of the below applications and packages installed.

## R with RStudio {#setup-r}

One of the most respected introductions to R is [R for Data Science](https://r4ds.had.co.nz/introduction.html) by Hadley Wickham and Garrett Grolemund, and the [Prerequisities](https://r4ds.had.co.nz/introduction.html#prerequisites) section is set up for installing R. 

## Common packages

We will use several common packages; if you haven't installed them yet, feel free to install them into R:

```{r setup install, eval = FALSE}
install.packages(c('tidyverse', 'lubridate', 'readr',
                   'DBI', 'devtools',
                   'gt', 'pryr', 'stringi',
                   'rmarkdown',
                   'tidymodels', 'tidytext', 'tm', 'rpart', 'rpart.plot',
                   'jsonlite', 'httr'))
```

## Highbond R package

For those customers who use the cloud-based Galvanize Highbond (and specifically the Results Module), we will use the R package [galvanizer](https://github.com/jonlinca/galvanizer) to use Results information in the Applied Analytics chapter. This package is maintained by Jonathan Lin during his spare time, and any issues or suggestions on its improvement would be appreciated.

```{r install github, eval = FALSE}
install.packages('galvanizer')
```


